Mergeable Types for Highly-Available Replicated Data

### 2017-02-23 paper/semantics.tex
Semantics.

\begin{figure}[!t]
\centering
\subcaptionbox[] {\small
  In this example, 1 and 3 have two LCAs (3 and 4) a result of
  previous merges. The dotted circle denotes a virtual ancestor
  obtained by merging the two LCAs.
  \label{fig:criss-cross-lcas}
} [0.47\columnwidth] {
  \includegraphics[scale=0.55]{Figures/2-LCAs}
}
\hfill
\subcaptionbox[] {\small
  In this example, 3 and 6 have two LCAs (2 and 1) despite there not
  being any previous merges between their respective branches.
  \label{fig:external-lcas}
} [0.47\columnwidth] {
  \includegraphics[scale=0.7]{Figures/2-external-LCAs}
}
\caption{Examples where merging versions have more than one LCA}
\label{fig:many-lcas}
\end{figure}

When merging two concurrent versions $v_1$ and $v_2$, the common
ancestor argument for \C{merge} must the LCA of $v_1$ and $v_2$,
without which \C{merge} may yeild unexpected results. This is
demonstrated for the grow-only counter in
Fig.~\ref{fig:merge-needs-lca}, where incorrect count is obtained if a
common ancestor that is not an LCA is used to merge 4 and 7. While in
this example there is a unique LCA for 4 and 7, in general this may
not be the case. With unrestrained branching and merging, there is no
bound on the number of LCAs a pair of versions can have.
For example, in
Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded by
two ``criss-cross'' merges between their respective branches resulting
in there being two LCAs (5 and 4) for 0 and 3. Multiple LCAs can occur
even without criss-cross merges, as demonstrated by
Fig.~\ref{fig:external-lcas}. The problem with concurrent versions
with multiple LCAs is that they do not lend themselves to three-way
merging. This problem also arises in the context of source
control systems, and often solved via ad-hoc mechanisms.
GitHub~\cite{github}, for instance, recursively merges LCAs to compute
a virtual ancestor, which then serves as the LCA for merging the
concurrent versions. This method is demonstrated for the example in
Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and 4 of 0 and 3 are
merged (with LCA of 10) to generate -1 as the virtual LCA to merge 5
and 4. A major downside with this method is that it makes no
guarantees of the relationship between the virtual ancestor and the
concurrent versions; the former may not even be a legal ancestor of
the latter as per the semantics of the data type. For instance, the
integer type in Fig.~\ref{fig:criss-cross-lcas} may in fact represent
a bank account, which disallows any activity on the account if balance
is ever known to be less than zero. Thus, from the perspective of a
bank account, it doesn't make sense how concurrent versions 3 and 0
emerged from -1, since the only transition allowed by the semantics
from -1 is to itself. Clearly, ad hoc mechanisms, which work for the
text, may not work for more sophisticated data types.

@@

Let us see how the aforementioned restriction allows us to merge the
branches in Fig.~\ref{fig:external-lcas} without ever creating
multiple LCAs. The legal branching history is shown in
Fig.~\ref{fig:legal-merge}. First, like in
Fig.~\ref{fig:external-lcas}, $b_2$ is merged into $b_1$ to create
$v_{12}$, and $b_3$ is merged into $b_4$ to create $v_{42}$. Thus, the
external locus of $b_1$ is $v_{22}$, and that of $b_4$ is $v_{32}$.
Observe that now $b_1$ cannot merge $b_3$, neither can $b_4$ merge
$b_2$ because $v_{22}$ and $v_{32}$ are not related by the ancestor
relation. Same applies for $b_1$ and $b_4$ because neither one's
latest version follows from other's external locus.  However, $b_2$
and $b_3$ can merge among themselves. Let us say they indeed merge to
to create a version $v_{23}$ on $b_2$. The branch $b_2$ is now
eligible to be merged into $b_1$ and also $b_4$. These merges do occur
in Fig.~\ref{fig:legal-merge} to create versions $v_{13}$ and
$v_{43}$, and making $v_{23}$ the external locus for both $b_1$ and
$b_4$. Thus, $b_1$ can now merge into $b_4$ to create version $v_{44}$
on $b_4$.

>>>>

\begin{figure}[!t]
\centering
\subcaptionbox[] {\small
  In this example, 1 and 3 have two LCAs (3 and 4) a result of
  previous merges. The dotted circle denotes a virtual ancestor
  obtained by merging the two LCAs.
  \label{fig:criss-cross-lcas}
} [0.47\columnwidth] {
  \includegraphics[scale=0.55]{Figures/2-LCAs}
}
\hfill
\subcaptionbox[] {\small
  In this example, versions $v_{13}$ and $v_{44}$ have two LCAs
  ($v_{22}$ and $v_{32}$)  despite there not being any previous merges
  between their respective branches.
  \label{fig:external-lcas}
} [0.47\columnwidth] {
  \includegraphics[scale=0.7]{Figures/2-external-LCAs}
}
\caption{Examples where merging versions have more than one LCA}
\label{fig:many-lcas}
\end{figure}

When merging two concurrent versions $v_1$ and $v_2$, the common
ancestor argument for \C{merge} must the LCA of $v_1$ and $v_2$,
without which \C{merge} may yeild unexpected results. This is
demonstrated for the grow-only counter in
Fig.~\ref{fig:merge-needs-lca}, where incorrect count is obtained if a
common ancestor that is not an LCA is used to merge 4 and 7. While in
this example there is a unique LCA for 4 and 7, in general this may
not be the case. With unrestrained branching and merging, there is no
bound on the number of LCAs a pair of versions can have.
For example, in
Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded by
two ``criss-cross'' merges between their respective branches resulting
in there being two LCAs (5 and 4) for 0 and 3. Multiple LCAs can occur
even without criss-cross merges, as demonstrated by
Fig.~\ref{fig:external-lcas}. Concurrent versions with multiple LCAs
do not lend themselves to three-way merging. This problem also arises
in the context of source control systems, which employ ad hoc
mechanisms to pave way for three-way merging.  GitHub~\cite{github},
for instance, recursively merges LCAs to compute a virtual ancestor,
which then serves as the LCA for merging the concurrent versions. This
method is demonstrated for the example in
Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and 4 of 0 and 3 are
merged (with LCA of 10) to generate -1 as the virtual LCA to merge 5
and 4. A major downside with this method is that it makes no
guarantees of the relationship between the virtual ancestor and the
concurrent versions; the former may not even be a legal ancestor of
the latter as per the semantics of the data type. For instance, the
integer type in Fig.~\ref{fig:criss-cross-lcas} may in fact represent
a bank account, which disallows any activity on the account if balance
is ever known to be less than zero. Thus, from the perspective of a
bank account, it doesn't make sense how concurrent versions 3 and 0
emerged from -1, since the only transition allowed by the semantics
from -1 is to itself. Clearly, ad hoc mechanisms, which work for the
text, may not work for more sophisticated data types.

@@

\begin{figure}
\centering
\includegraphics[scale=0.75]{Figures/legal-merge}

\caption{A legal branch history for the example in
Fig.~\ref{fig:external-lcas}}
\label{fig:legal-merge}
\end{figure}

Let us see how the aforementioned restriction allows us to merge the
branches in Fig.~\ref{fig:external-lcas} without ever creating
multiple LCAs. The legal branching history is shown in
Fig.~\ref{fig:legal-merge}. First, like in
Fig.~\ref{fig:external-lcas}, $b_2$ is merged into $b_1$ to create
$v_{12}$, and $b_3$ is merged into $b_4$ to create $v_{42}$. Thus, the
external locus of $b_1$ is $v_{22}$, and that of $b_4$ is $v_{32}$.
Observe that now $b_1$ cannot merge $b_3$, neither can $b_4$ merge
$b_2$ because $v_{22}$ and $v_{32}$ are not related by the ancestor
relation. Same applies for $b_1$ and $b_4$ because neither one's
latest version follows from other's external locus.  However, $b_2$
and $b_3$ can merge among themselves. Let us say they indeed merge to
to create a version $v_{23}$ on $b_2$. The branch $b_2$ is now
eligible to be merged into $b_1$ and also $b_4$. These merges do occur
in Fig.~\ref{fig:legal-merge} to create versions $v_{13}$ and
$v_{44}$, and making $v_{23}$ the external locus for both $b_1$ and
$b_4$. Thus, $b_1$ can now merge into $b_4$ to create version $v_{45}$
on $b_4$. Note that alternative legal histories are also possible. For
instance, $b_4$ could have skipped the merge with $v_{23}$ (hence, the
version $v_{44}$), and directly merged $v_{13}$ in the last step.
This is legal because $b_4$'s locus before the merge, $v_32$, is an
ancestor of the merging version $v_{13}$.
